---
title: "Quick Start Guide"
subtitle: "Get up and running with bayesTPC in minutes"
format: html
whitespace: small
---

This quick start guide will get you fitting thermal performance curves with `bayesTPC` in just a few minutes. For detailed tutorials with real datasets and advanced techniques, see our [Tutorials](materials.qmd) page.

# Installation

First, install the required packages:

```{r eval=FALSE}
# Install nimble first
install.packages("nimble")

# Install bayesTPC from GitHub
remotes::install_github("johnwilliamsmithjr/bayesTPC")
```

# Load packages

```{r}
library(nimble)
library(bayesTPC)
```

# Your first TPC fit

## 1. Prepare your data

Your data should be in a list with two elements: `Trait` (your response variable) and `Temp` (temperature in °C).

```{r}
# Example: development rate data
data <- read.csv("data/VByte_106_development_rate.csv")
data_list <- list(Trait = data$OriginalTraitValue, Temp = data$Interactor1Temp)

# Quick plot to see the data
plot(data_list$Trait ~ data_list$Temp, ylab = "Development Rate", xlab = "Temperature (°C)", pch = 16)
```

## 2. Choose a model

`bayesTPC` comes with many pre-built models. See what's available:

```{r}
get_models()
```

For this example, we'll use the Briere model, which is good for asymmetric unimodal curves.

## 3. Fit the model

```{r}
set.seed(1)
fit <- b_TPC(data = data_list, model = "briere")
```

## 4. Check the results

```{r}
# Basic summary
print(fit)

# Plot the fit
plot(fit, temp_interval = seq(0, 35, length.out = 1000))
```

That's it! You've successfully fit a thermal performance curve.

# Next steps

This quick start covered the basics. For more detailed tutorials, check out:

- **[Basic Tutorial](EEID2024_basic.qmd#basic-packages)**: Complete workflow with diagnostics, model selection, and best practices
- **[Advanced Tutorial](EEID2024_advanced.qmd#adv-flexTPC)**: Custom model creation and advanced analysis techniques
- **[VectorByte Workshop](VB_Bayes_activity2B.qmd)**: Hands-on activity with mosquito trait data

# Common issues and solutions

## Poor convergence
If your model doesn't converge well, try:
- Adjusting priors: `priors = list(T_min = "dunif(-5,10)", T_max = "dunif(30,40)")`
- Using a different sampler: `samplerType = "AF_slice"`
- Adding initial values: `inits = list(T_min = 10, T_max = 32)`

## Model selection
Compare models using WAIC:
```{r eval=FALSE}
fit1 <- b_TPC(data_list, "briere")
fit2 <- b_TPC(data_list, "quadratic")
bayesTPC::get_WAIC(fit1)
bayesTPC::get_WAIC(fit2)
```

## Custom models
Create your own TPC function:
```{r eval=FALSE}
my_model <- specify_normal_model("my_model", 
                                parameters = c(q = "dunif(0,1)", T_max = "dunif(30,50)"),
                                formula = expression(q * Temp * (T_max - Temp)))
```

For detailed explanations and examples, see our [full tutorials](materials.qmd).