---
title: "Getting Started with bayesTPC"
subtitle: "Learn to fit thermal performance curves in minutes"
format: html
whitespace: small
---

# Welcome to bayesTPC

This guide will get you fitting thermal performance curves (TPCs) with `bayesTPC` quickly and efficiently. Whether you're new to Bayesian methods or an experienced user, we have tutorials designed for your skill level.

## Quick Start (5 minutes)

### Installation

First, install the required packages:

```{r eval=FALSE}
# Install nimble first
install.packages("nimble")

# Install bayesTPC from GitHub
remotes::install_github("johnwilliamsmithjr/bayesTPC")
```

### Your First TPC Fit

```{r}
library(nimble)
library(bayesTPC)

# Load example data
data <- read.csv("data/AeaegyptiTraitData.csv")
data_list <- list(Trait = data$trait, Temp = data$Temp)

# Quick plot
plot(data_list$Trait ~ data_list$Temp, ylab = "Development Rate", xlab = "Temperature (°C)", pch = 16)
```

```{r}
# Choose a model and fit
set.seed(1)
fit <- b_TPC(data = data_list, model = "briere")

# Check results
print(fit)
plot(fit, temp_interval = seq(0, 35, length.out = 1000))
```

That's it! You've successfully fit a thermal performance curve.

## Tutorial Series

### For Beginners: Complete Workflow

**[EEID 2024 Basic Tutorial](EEID2024_basic.qmd)** - Perfect for new users
- **Dataset**: *Aedes aegypti* mosquito traits (development rate, adult longevity, juvenile survival)
- **What you'll learn**: Data formatting, model fitting, diagnostics, model comparison
- **Time**: 30-45 minutes
- **Jump to**: [Packages](EEID2024_basic.qmd#basic-packages) | [Data setup](EEID2024_basic.qmd#basic-aedes-data) | [Examples](EEID2024_basic.qmd#basic-example1) | [Practice](EEID2024_basic.qmd#basic-practice)

### For Advanced Users: Custom Models

**[EEID 2024 Advanced Tutorial](EEID2024_advanced.qmd)** - For experienced users
- **Dataset**: *E. coli* growth under antibiotics
- **What you'll learn**: Custom model creation, posterior transformations, advanced analysis
- **Time**: 45-60 minutes
- **Jump to**: [Custom models](EEID2024_advanced.qmd#adv-flexTPC) | [Fitting](EEID2024_advanced.qmd#adv-fitting) | [Transformations](EEID2024_advanced.qmd#adv-transformations)

### Hands-on Practice

**[VectorByte Workshop](VB_Bayes_activity2B.qmd)** - Interactive activity
- **Focus**: Mosquito trait modeling
- **Format**: Self-contained exercise
- **Time**: 20-30 minutes

## Learning Paths

### **New to bayesTPC?**
1. **Quick Start** (above) - 5 minutes
2. **[Basic Tutorial](EEID2024_basic.qmd#basic-packages)** - Complete workflow
3. **[VectorByte Workshop](VB_Bayes_activity2B.qmd)** - Hands-on practice
4. **[Advanced Tutorial](EEID2024_advanced.qmd#adv-flexTPC)** - Custom models

### **Experienced user?**
- Jump directly to **[Advanced Tutorial](EEID2024_advanced.qmd#adv-flexTPC)** for custom model creation
- Check **[Model comparison](EEID2024_advanced.qmd#adv-model-comparison)** for WAIC analysis
- Explore **[Posterior transformations](EEID2024_advanced.qmd#adv-transformations)** for advanced analysis

## Available Datasets

All tutorials use real biological datasets:

- **[VecTraits Database](https://vectorbyte.crc.nd.edu/vectraits-explorer)**: Access thousands of vector trait datasets
- **[Antibiotic resistance data](data/ab_data.csv)**: *E. coli* growth under various antibiotics
- **[Mosquito trait data](data/AeaegyptiTraitData.csv)**: Development, survival, and longevity data

## Common Issues & Solutions

### Poor convergence?
Try these adjustments:
```{r eval=FALSE}
# Adjust priors
priors = list(T_min = "dunif(-5,10)", T_max = "dunif(30,40)")

# Use different sampler
samplerType = "AF_slice"

# Add initial values
inits = list(T_min = 10, T_max = 32)
```

### Model selection?
Compare models using WAIC:
```{r eval=FALSE}
fit1 <- b_TPC(data_list, "briere")
fit2 <- b_TPC(data_list, "quadratic")
bayesTPC::get_WAIC(fit1)
bayesTPC::get_WAIC(fit2)
```

### Custom models?
Create your own TPC function:
```{r eval=FALSE}
my_model <- specify_normal_model("my_model", 
                                parameters = c(q = "dunif(0,1)", T_max = "dunif(30,50)"),
                                formula = expression(q * Temp * (T_max - Temp)))
```

## Need Help?

- **Diagnostics**: See [troubleshooting section](EEID2024_basic.qmd#basic-example1) in Basic Tutorial
- **Convergence**: Check [convergence tips](EEID2024_advanced.qmd#adv-fitting) in Advanced Tutorial
- **Package issues**: Ensure you're using `nimble` version 0.13.2 (see [README](README.md))

## What's Next?

Once you're comfortable with the basics, explore:
- **Different TPC models**: `get_models()` shows all available options
- **Custom likelihoods**: Beyond normal distributions
- **Model comparison**: Using WAIC and other criteria
- **Real data**: Apply to your own research questions

Ready to dive deeper? Start with the **[Basic Tutorial](EEID2024_basic.qmd#basic-packages)** for a comprehensive workflow!